Differential Protein expression

### 2021-11-12 reports/rmarkdown/manuscript/Figure2.Rmd

---
title: "Main Figure 2. Bacterial Communities and Function"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes: 
- \usepackage{float}
editor_options: 
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
##################
# LOAD LIBRARIES #
##################
#setwd("/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/reports/rmarkdown/manuscript")
suppressWarnings({suppressMessages({suppressPackageStartupMessages({
  library(tidyverse)
  library(edgeR)
  library(openxlsx)
  library(scales)
  library(igraph)
  library(RColorBrewer)
  remotes::install_github("czarnewski/niceRplots",force=T)
  library(niceRplots)
})  })  })

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8}

layout(matrix(c(1,1,2,2,2,
                1,1,3,3,3,
                1,1,4,4,4,
                5,6,7,8,9,
                5,6,7,8,10),
              nrow = 5,ncol = 5,byrow = T), widths = c(.6,1.5,.2,1.9,2),
       heights = c(.6,.6,.6,1,1.5))
figlabels <- letters

## A
#########################
# BACTERIAL COMMUNITIES #
#########################
bac_communities <- factor(paste0("BC",sprintf("%02d",bac_communities)))
gB <- graph_from_adjacency_matrix(as.matrix(SNN_bacteria), mode = "undirected",diag = F,weighted = T)
 set.seed(1)
lB <- layout_nicely(gB,niter=3000,start.temp=30)

par(mar=c(0,0,3.5,5))
plot( gB , vertex.label.cex=0.000001 , vertex.color = bact_pal[factor(bac_communities)] ,
      edge.width=  ( E(gB)$weight / max(E(gB)$weight)) ,
      vertex.size=10,
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(gB)$weight / max(E(gB)$weight) * 88 )+1 ] ,layout=lB)
title(main = "           Bacterial Communities- Louvain", line = 0.5, cex.main = 1)
legend(par("usr")[2],par("usr")[4],title.adj = 0,
       legend = levels(factor(bac_communities)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = bact_pal,pt.cex = 1,xpd=T)

@@

## B
###################################
# BACTERIAL VAGINOSIS COMPARISSON #
###################################
par(mar=c(1,5,2,1)) #b,l,t,r
x <- datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]]
xL <- rowsum(x , grepl("L.",rownames(x)) )[2,] / sum(grepl("L.",rownames(x)))
xM <- rowsum(x , grepl("Mobiluncus",rownames(x)) )[2,] / sum(grepl("Mobiluncus",rownames(x)))
xG <- rowsum(x , grepl("Gardnerella",rownames(x)) )[2,] / sum(grepl("Gardnerella",rownames(x)))

@@

barlist( data = rbind(WS=mL ,
                      "16S counts"=xL),
         main = "Lactobacillus",
         genes = c("WS","16S counts"),
         clustering = group_annotation,
         draw_mean_lines=F,col = pal,srt=20)

@@

barlist( data = rbind(WS=mM ,
                      "16S counts"=xM),
         main = "Mobiluncus",
         genes = c("WS","16S counts"),
         clustering = group_annotation,
         draw_mean_lines=F,col = pal,srt=20)

barlist( data = rbind(WS=mG ,
                      "16S counts"=xG),
         main = "Gardnerella",
         genes = c("WS","16S counts"),
         clustering = group_annotation,
         draw_mean_lines=F,col = pal,srt=20)

## C
#####################
# BACTERIAL PICRUST #
#####################
picrust_data <- round(picrust_data[ rowSums(picrust_data>5)>=10, ])
TRX <- datasets_all_samples[["Tissue_RNAseq_V3_normalized"]]
sample_use <- colnames(TRX)[colSums(TRX)!=0]

@@

named_picrust_data <- rowsum( picrust_data , group = ko_names[ rownames(picrust_data) ] )
#named_picrust_data[1:20,1:3]
#dim(named_picrust_data)
named_picrust_data <- round(named_picrust_data[ rowSums(named_picrust_data>5)>=5, ])

@@

par(mar=c(4,.1,.5,.2)) #b,l,t,r

h <- hclust( as.dist( (1- cor(t(top_picrust_data)))/2 ), method = "ward.D2")
plot( rev(as.dendrogram(h)) ,xlim=c(max(h$height),-1), horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=T)
cutoff <- 4
abline(v=cutoff,xpd=F,col="red",lty=2)
gene_module <- cutree(h, h = cutoff)

@@

image( t(top_picrust_data[h$order,][nrow(top_picrust_data):1,order(group_annotation[sample_use])]),
       col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
       breaks = seq(-5,5,length.out = 100),axes=F)
points( seq(0,1,length.out = length(sample_use)),
        rep(par("usr")[4],length(sample_use))+.005, pch=15, cex=.7,xpd=T,
        col=pal[factor(group_annotation[sample_use])[ order(group_annotation[sample_use])]])

points( seq(0,1,length.out = ncol(top_picrust_data)), 
        rep(par("usr")[3],ncol(top_picrust_data))-.008, pch=15, xpd=T,cex=.5,
        col= c("tomato","orange","#d7d7d7")[as.factor(metadata$BV_Diagnosis_v3)][order( group_annotation[sample_use])] )

points( seq(0,1,length.out = ncol(top_picrust_data)), 
        rep(par("usr")[3],ncol(top_picrust_data))-.024, pch=15, xpd=T,cex=.5,
        col= c("#d7d7d7","tomato")[as.factor(metadata$HIVstatus)][order(group_annotation[sample_use])])

@@

par(mar=c(4,.1,.5,7))
barplot( rev(pvalues) , horiz = T , col= taxa_pal[factor(rev(module_color))],
         xaxs="i",yaxs="i",xlim=c(0,1.2*max(pvalues)),ylab="",las=1,names.arg = "")
title(xlab="-log10(pvalue)", line=2)
abline(v=2,xpd=F,col="red",lty=2)
points( rev(pvalues), seq(1,length(pvalues) )*1.2-.5,pch=21,bg=taxa_pal[factor(rev(module_color))] )
text( rev(pvalues)+strwidth("M") , seq(1,length(pvalues) )*1.2-.5, adj=0,cex=1,xpd=T,
      labels = rev(gsub("\\(\\)","",names(pvalues)) ),xpd=T)
text( rev(pvalues)+strwidth("M") , seq(1,length(pvalues) )*1.2-.5, adj=0,cex=1,
      labels = rev(names(pvalues)),col = paste0(taxa_pal[factor(rev(module_color))],90) ,xpd=T)

## D
################
# PICRUST UMAP #
################

>>>>

---
title: "Figure 2. Bacterial Communities and Function"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes: 
- \usepackage{float}
editor_options: 
  chunk_output_type: console
---

```{r Load data, message=FALSE, warning=FALSE, include=FALSE}
##################
# LOAD LIBRARIES #
##################
#setwd("/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/reports/rmarkdown/manuscript")
suppressWarnings({suppressMessages({suppressPackageStartupMessages({
  library(tidyverse)
  library(edgeR)
  library(openxlsx)
  library(scales)
  library(igraph)
  library(RColorBrewer)
  remotes::install_github("czarnewski/niceRplots",force=T)
  library(niceRplots)
})  })  })

@@

```{r Figure 2.,echo=FALSE, message=FALSE, warning=FALSE, fig.height=8}

layout(matrix(c(1,1,2,2,2,
                1,1,3,3,3,
                1,1,4,4,4,
                5,6,7,8,9,
                5,6,7,8,10),
              nrow = 5,ncol = 5,byrow = T), widths = c(.6,1.5,.2,1.9,2),
       heights = c(.68,.56,.56,1,1.5))
figlabels <- letters

### A
#########################
# BACTERIAL COMMUNITIES #
#########################
bac_communities <- factor(paste0("BC",sprintf("%02d",bac_communities)))
gB <- graph_from_adjacency_matrix(as.matrix(SNN_bacteria), mode = "undirected",diag = F,weighted = T)
 set.seed(1)
lB <- layout_nicely(gB,niter=3000,start.temp=30)

par(mar=c(0,0,3.5,5))
plot( gB , vertex.label.cex=0.000001 , vertex.color = bact_pal[factor(bac_communities)] ,
      edge.width=  ( E(gB)$weight / max(E(gB)$weight)) ,
      vertex.size=10,
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(gB)$weight / max(E(gB)$weight) * 88 )+1 ] ,layout=lB)
title(main = "           Bacterial Communities- Louvain", line = 1, cex.main = 1)
legend(par("usr")[2],par("usr")[4],title.adj = 0,
       legend = levels(factor(bac_communities)),xjust = 0,yjust = 1,
       bty = "n",pch = 21,pt.bg = bact_pal,pt.cex = 1,xpd=T)

@@

### B
###################################
# WETT SMEAR BACTERIAL VALIDATION #
###################################
par(mar=c(1,5,4,1)) #b,l,t,r
x <- datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]]
xL <- rowsum(x , grepl("L.",rownames(x)) )[2,] / sum(grepl("L.",rownames(x)))
xM <- rowsum(x , grepl("Mobiluncus",rownames(x)) )[2,] / sum(grepl("Mobiluncus",rownames(x)))
xG <- rowsum(x , grepl("Gardnerella",rownames(x)) )[2,] / sum(grepl("Gardnerella",rownames(x)))

@@

barlist( data = rbind(WS=mL ,
                      "16S counts"=xL),
         genes = c("WS","16S counts"),
         clustering = group_annotation,srt=20,
         draw_mean_lines=F,col = pal)
mtext("Lactobacillus", side=3, las = 1, cex = .6, line = .5,xpd=T)
mtext("Wett Smear Bacteria Validation", side=3, las = 1, cex = .6, line =1.5,xpd=T, font = 2)

@@

par(mar=c(1,5,2,1)) #b,l,t,r
barlist( data = rbind(WS=mM ,
                      "16S counts"=xM),
         genes = c("WS","16S counts"),
         clustering = group_annotation,srt=20,
         draw_mean_lines=F,col = pal)
mtext("Mobiluncus", side=3, las = 1, cex = .6, line = .5,xpd=T)

barlist( data = rbind(WS=mG ,
                      "16S counts"=xG),
         genes = c("WS","16S counts"),
         clustering = group_annotation,srt=20,
         draw_mean_lines=F,col = pal)
mtext("Gardnerella", side=3, las = 1, cex = .6, line = .5,xpd=T)

### C
#####################
# BACTERIAL PICRUST #
#####################
picrust_data <- round(picrust_data[ rowSums(picrust_data>5)>=10, ])
TRX <- datasets_all_samples[["Tissue_RNAseq_V3_normalized"]]
sample_use <- intersect(colnames(TRX)[colSums(TRX)!=0], colnames(picrust_data))

@@

named_picrust_data <- rowsum( picrust_data , group = ko_names[ rownames(picrust_data) ] )
named_picrust_data <- round(named_picrust_data[ rowSums(named_picrust_data>5)>=5, ])

@@

par(mar=c(4,.1,4,.2)) #b,l,t,r
h <- hclust( as.dist( (1- cor(t(top_picrust_data)))/2 ), method = "ward.D2")
plot( rev(as.dendrogram(h)) ,xlim=c(max(h$height),-1), horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=T)
title(xlab="Height", line=2)
cutoff <- 4
abline(v=cutoff,xpd=F,col="red",lty=2)
gene_module <- cutree(h, h = cutoff)

@@

image( t(top_picrust_data[h$order,][nrow(top_picrust_data):1,order(group_annotation[sample_use])]),
       col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
       breaks = seq(-5,5,length.out = 100),axes=F)
mtext("Differentialy Expressed KEGG Ontology Terms", side=3, las = 1, cex = .6, line =1.5,xpd=T, font = 2)
title(xlab="Study Groups", line=1)

end <- table(pal[factor(group_annotation)[ order(group_annotation[sample_use])]] )
end <- end[order(factor(names(end), levels = pal))]
end <- map_dbl(cumsum(end), ~ (.x * (par("usr")[2])/ncol(picrust_data[sample_use])-.015) )
start <- c(0, map_dbl(end[1:4], ~.x+0.015))

axis(1, at = c(start, end), label = F, pos =1.027 , xpd=T, col="white",col.ticks="black")
map2(start, end, ~lines(x=c(.y, .x), y= c(1.027, 1.027), xpd=T, cex=.8))
# text labels
t <- map2(start, end, ~(.x+.y)/2)
map2(t,levels(factor(group_annotation)), ~text(x=.x, y=1.043, .y,xpd=T, cex=.8))

points( seq(0,1,length.out = length(sample_use)),
        rep(par("usr")[4],length(sample_use))+.005, pch=15, cex=.7,xpd=T,
        col=pal[factor(group_annotation[sample_use])[ order(group_annotation[sample_use])]])
points( seq(0,1,length.out = ncol(top_picrust_data)), 
        rep(par("usr")[3],ncol(top_picrust_data))-.008, pch=15, xpd=T,cex=.5,
        col= c("tomato","orange","#d7d7d7")[as.factor(metadata$BV_Diagnosis_v3)][order( group_annotation[sample_use])] )
points( seq(0,1,length.out = ncol(top_picrust_data)), 
        rep(par("usr")[3],ncol(top_picrust_data))-.024, pch=15, xpd=T,cex=.5,
        col= c("#d7d7d7","tomato")[as.factor(metadata$HIVstatus)][order(group_annotation[sample_use])])

@@

par(mar=c(4,.1,4,7))
barplot( rev(pvalues) , horiz = T , col= taxa_pal[factor(rev(module_color))],
         xaxs="i",yaxs="i",xlim=c(0,1.2*max(pvalues)),ylab="",las=1,names.arg = "")
title(xlab="-log10(p-value)", line=2)
abline(v=2,xpd=F,col="red",lty=2)
points( rev(pvalues), seq(1,length(pvalues) )*1.2-.5,pch=21,bg=taxa_pal[factor(rev(module_color))] )
text( rev(pvalues)+strwidth("M") , seq(1,length(pvalues) )*1.2-.5, adj=0,cex=1,xpd=T,
      labels = rev(gsub("\\(\\)","",names(pvalues)) ),xpd=T)
text( rev(pvalues)+strwidth("M") , seq(1,length(pvalues) )*1.2-.5, adj=0,cex=1,
      labels = rev(names(pvalues)),col = paste0(taxa_pal[factor(rev(module_color))],90) ,xpd=T)

### D
################
# PICRUST UMAP #
################

### 2021-11-12 reports/rmarkdown/manuscript/Figure8.Rmd

# save TopTables
write.xlsx(dpe_L$top, file=paste0("../../../results/DE/","dpe_across_","Luminal",dpe_L$c,".xlsx"))
write.xlsx(dpe_T$top, file=paste0("../../../results/DE/","dpe_across_","Tissue",dpe_T$c,".xlsx"))

@@

# save TopTables 
l_top <- list(Luminal = pairwise_dpe_L$top, Tissue = pairwise_dge_T$top)
imap(l_top, ~write.xlsx(.x,file=paste0("../../../results/DE/","dpe_pairwise_",.y,"_conf",".xlsx"))) # Excel

@@

```{r heatmap functions, echo=FALSE}

top_dpe.fun <- function(top){
  p_val <- pull(top, "P.Value")
  top_dpe <- (p_val < 0.05) #& (rowSums( abs(top[,grep("logFC",colnames(top))]) >= log2(1.5) ) >= 1)
  top_dpe <- top$Antibody[ top_dpe ]

@@

stat <- imap(c("Luminal_gr", "Tissue_gr"), ~test.krusk_dunn.fun(cytogr.long, Cytokines, .x))
s <- flatten(stat)
write.xlsx(s, file=paste0("../../../results/","Cytokine_stats",".xlsx"))

@@

**Figure 3. Protein Heatmap**. **a** and **c** Without scaling (normalized MFI), for Luminal and Tissue respectively **b** and **d** scaled by row for Luminal and Tissue respectively. top dpe with p-value < 0.05

>>>>

@@

@@

```{r heatmap functions, echo=FALSE}
top_dpe.fun <- function(top){
  p_val <- pull(top, "P.Value")
  top_dpe <- (p_val < 0.05) #& (rowSums( abs(top[,grep("logFC",colnames(top))]) >= log2(1.5) ) >= 1)
  top_dpe <- top$Antibody[ top_dpe ]

@@

n <- c("Luminal_gr", "Tissue_gr") %>% set_names(c("Luminal", "Tissue"))
stat <- imap(n, ~test.krusk_dunn.fun(cytogr.long, Cytokines, .x))
s <- flatten(stat)
```

```{r Save stats to excel, include=FALSE}
# save TopTables pairwise
l_top_p <- list(Luminal = pairwise_dpe_L$top, Tissue = pairwise_dge_T$top)
imap(l_top_p, ~write.xlsx(.x,file=paste0("../../../results/DE/","dpe_pairwise_",.y,".xlsx"))) # Excel

# save TopTables across + cytokines
l_top_a <- list(Luminal = list(Across = dpe_L$top), Tissue = list(Across = dpe_T$top))
l <- list(l_top_a, stat, l_top_p)
names <- c("Luminal", "Tissue") %>% set_names()
p_list <- imap(names, ~flatten(map_depth(l, 1, .x)))

imap(p_list, ~write.xlsx(.x,file=paste0("../../../results/DE/","Proteins_",.y,".xlsx"))) # Excel

@@

**Figure 3. Protein Heatmap**. **a** and **b** top differentially expressed proteins scaled by row, for Luminal and Tissue respectively. (p-value < 0.05) **b** and **d** sig. cytokines for Luminal and Tissue respectively.
